---
title: "Birthday_investigation"
output: html_document
editor_options: 
  chunk_output_type: console
---

# check working directory
```{r}
getwd()
setwd("/Users/geodev/code/daan-nano/term2/EDA/lesson4/")
list.files()
```

# import packages
```{r}
install.packages("tidyr")
```

```{r}
install.packages("dplyr")
```

```{r}
install.packages("lubridate")
```

# read libraries
```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(lubridate)
```

# read data into dataframe
```{r}
# define path name
inputFileName <- "birthdaysExample.csv"
bdays_raw <- read.csv(inputFileName)
head(bdays_raw)
```

# data quicklook
```{r}
str(bdays_raw)
names(bdays_raw)
typeof(bdays_raw)
```

# convert strings to datetime
```{r}
#bdays_edt <- cbind(bdays_raw, datesFormatted = as.Date(bdays_raw$dates, "%m/%d/%Y"))
bdays_edt <- cbind(bdays_raw, datesFormatted = as.POSIXct(bdays_raw$dates, tz="", format="%m/%d/%Y"))
str(bdays_edt)
```

# create by_week values
```{r}
date_by_week = floor_date(bdays_edt$datesFormatted, unit="week")
```

# add date_by_week values to df
```{r}
bdays_edt <- cbind(bdays_edt, date_by_week = date_by_week)
```

# count by week
```{r}
bdays_per_week <- count(bdays_edt, vars=date_by_week)
bdays_per_week
```

# create dates_by_month values
```{r}
date_by_month = floor_date(bdays_edt$datesFormatted, unit = "month")
```

# add date_by_month values to df
```{r}
bdays_edt <- cbind(bdays_edt, date_by_month =date_by_month)
```

# count by month
```{r}
bdays_per_month <- count(bdays_edt, vars=date_by_month)
```

# create dates_by_day values
```{r}
date_by_day = floor_date(bdays_edt$datesFormatted, unit = "day")
```

# add date_by_day values to df
```{r}
bdays_edt <- cbind(bdays_edt, date_by_day =date_by_day)
```

# count by day
```{r}
bdays_per_day <- count(bdays_edt, vars=date_by_day)
```

# max values
```{r}
max(bdays_per_day$n)
which.max(bdays_per_day$n)
max(bdays_per_week$n)
which.max(bdays_per_week$n)
max(bdays_per_month$n)
which.max(bdays_per_month$n)
```

# count bdays
```{r}
count(bdays_edt)
```

# count events on given day
```{r}
sum(bdays_edt$datesFormatted == "0014-01-07")
```

# plot series 1 - simple histograms:
```{r}
# create plots
plot1 <- ggplot(bdays_edt, mapping = aes(bdays_edt$datesFormatted)) +
  geom_histogram() +
  #geom_freqpoly(binwidth = 0.05) +
  labs(x = "date", y = "number of birthdays", title="Histogram number of birthdays vs time")

plot2 <- ggplot(bdays_edt, mapping = aes(bdays_edt$date_by_month)) +
  geom_histogram() +
  #geom_freqpoly(binwidth = 0.05) +
  labs(x = "date", y = "number of birthdays", title="Histogram number of birthdays vs month")

plot3 <- ggplot(bdays_edt, mapping = aes(bdays_edt$date_by_week)) +
  geom_histogram() +
  #geom_freqpoly(binwidth = 0.05) +
  labs(x = "date", y = "number of birthdays", title="Histogram number of birthdays vs week")

# arrange plots in grid
grid.arrange(plot1, plot2, plot3, ncol=1)
```