---
title: "coalInvestigation"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Quicklook on Coal consumption, per person from:
# [gapminder](https://www.gapminder.org/data/)

# check working directory
```{r}
getwd()
setwd("/Users/geodev/code/daan-nano/term2/EDA/lesson4/")
list.files()
```

# import packages
```{r}
install.packages("tidyr")
```
```{r}
install.packages("dplyr")
```

# read libraries
```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(readxl)
library(gridExtra)
library(grid)
```

# read data into dataframe
```{r}
# define path name
fileName <- "CoalComsumptionPerCapita.xlsx"
coal_raw <- read_xlsx(fileName)
```

# remove rows with all NA values
```{r}
rowSums(is.na(coal_raw))
ncol(coal_raw)
coal_clean <- coal_raw[rowSums(is.na(coal_raw)) != (ncol(coal_raw) - 1),]
```

# check data set
```{r}
#summary(coal_clean)
str(coal_clean)
```

# years as integers
```{r}
#as.integer(x)
names(coal_clean)[-1] <- as.integer(names(coal_clean)[-1])
```

# rename first column
```{r}
colnames(coal_clean)[1] <- "country"
```

# gather() data
```{r}
# "Coal Consumption per person (tonnes oil equivalent)"
coal_clean <- gather(coal_clean, key = "year", value = "Coal_Consumption", 2:48)
```

# remove special characters
```{r}
coal_clean$country <- gsub(' ', '_', coal_clean$country)
coal_clean$country <- gsub(',', '', coal_clean$country)
```

# drop rows with na method
```{r}
# check if row has na
row_has_na <- apply(coal_clean, 1, function(x){any(is.na(x))})
# count rows with na
sum(row_has_na)
# drop rows with na
coal_clean <- coal_clean[!row_has_na,]
```

# set year as number
```{r}
#as.integer(x)
names(coal_clean)[2]
names(coal_clean)[2] <- as.integer(names(coal_clean)[2])
str(coal_clean)
```


# plot series 1 - simple histograms: these are not very legible
```{r}
# create plots
yr_plot <- ggplot(coal_clean, mapping = aes(coal_clean$Coal_Consumption, color=year)) +
  geom_histogram() +
  #geom_freqpoly(binwidth = 0.05) +
  labs(x = "Coal Consumption per person (tonnes oil equivalent)", y = "person count", title="Histogram of Coal Consumption per person and per country")

country_plot <- ggplot(coal_clean, mapping = aes(coal_clean$Coal_Consumption, color=country)) +
  geom_histogram() +
  labs(x = "Coal Consumption per person (tonnes oil equivalent)", y = "person count",title="Histogram of Coal Consumption per person and per country")

# arrange plots in grid
grid.arrange(yr_plot, country_plot, ncol=1)
```

# plot series 2 - boxplot
```{r}
# create plots
plot1 <- ggplot(coal_clean, aes(coal_clean$year, coal_clean$Coal_Consumption)) +
  geom_boxplot() +
  labs(x="year", y="Coal Consumption per person (tonnes oil equivalent)", title="Boxplots of Coal consumption vs Time")

# reorder boxplot by median THANKS AND CREDIT TO @Winston on http://bit.ly/2GMZAq8
plot2 <- ggplot(coal_clean, aes(x=reorder(coal_clean$country, coal_clean$Coal_Consumption, FUN = median), coal_clean$Coal_Consumption)) +
  geom_boxplot() +
  labs(x="Country", y="Coal Consumption per person (tonnes oil equivalent)", title="Boxplots of Coal consumption by country") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
# arrange plots in grid
grid.arrange(plot1, plot2, ncol=1)
```

# scatter plot: compare two countries based on their boxplots and geographical ties
# using this link:
# [carbonbrief.org](http://bit.ly/2LCPD2g)
# Indonesia is one of the World's top coal exporters and India is a large consumer of indonesian coal
# test1
```{r}
ccVSyr <- ggplot(aes(x = year, y = Coal_Consumption), data = coal_clean) + 
  geom_point()
ccVSctr <- ggplot(aes(x = country, y = Coal_Consumption), data = coal_clean) + 
  geom_point()
grid.arrange(ccVSyr, ccVSctr, ncol= 1)
```

# test2
```{r}
ggplot(data = subset(coal_clean, coal_clean$country == 'Indonesia' | coal_clean$country == 'India'), aes(x = year, y = Coal_Consumption)) + 
  geom_point()
```

# coal_clean$year as numeric
```{r}
coal_clean$year <- as.integer(coal_clean$year)
typeof(coal_clean$year)
```

# correlations
```{r}
year_CC_correl_indonesia <- with(subset(coal_clean, coal_clean$country == "Indonesia"), cor.test(year, Coal_Consumption, method = 'pearson'))
year_CC_correl_indonesia

year_CC_correl_india <- with(subset(coal_clean, coal_clean$country == "India"), cor.test(year, Coal_Consumption, method = 'pearson'))
year_CC_correl_india
```

# scatter plot
```{r}
ggplot(data = subset(coal_clean, coal_clean$country == 'Indonesia' | coal_clean$country == 'India'), aes(x = country, y = Coal_Consumption)) + 
  geom_point()
```

# clean data for 2 variable plots
# remove rows with all NA values
```{r}
rowSums(is.na(coal_raw))
ncol(coal_raw)
coal_clean2 <- coal_raw[rowSums(is.na(coal_raw)) != (ncol(coal_raw) - 1),]
```
# years as integers
```{r}
names(coal_clean2)[-1] <- as.integer(names(coal_clean2)[-1])
```
# rename first column
```{r}
colnames(coal_clean2)[1] <- "country"
```

# scatter plot
```{r}
yrPlot1 <- ggplot(data = coal_clean2, aes(x = coal_clean2$`1965`, y = coal_clean2$`1966`, colour=country)) + 
  geom_point() +
  labs(title="Year on year consumption", x=1965, y=1966)

yrPlot2 <- ggplot(data = coal_clean2, aes(x = coal_clean2$`1966`, y = coal_clean2$`1990`, colour=country)) + 
  geom_point() +
  labs(title="Year on year consumption", x=2010, y=2011)

grid.arrange(yrPlot1, yrPlot2, ncol=1)
```

# Consumption By Year
```{r}
consumptionByYear <- coal_clean %>% 
  group_by(year) %>% 
  summarise(mean_cons = mean(Coal_Consumption),
            median_cons = median(Coal_Consumption),
            min_cons = min(Coal_Consumption),
            max_cons = max(Coal_Consumption),
            n = n())

tail(consumptionByYear)
```

# bar charts
```{r}
p_mc_1 <- ggplot() +
  geom_line(data = consumptionByYear, aes(x = year, y = mean_cons), color = 'red') +
  geom_line(data = consumptionByYear, aes(x = year, y = median_cons), color = 'green') + 
  geom_line(data = consumptionByYear, aes(x = year, y = min_cons), color = 'blue') +
  geom_line(data = consumptionByYear, aes(x = year, y = max_cons), color = 'black') +
  ylab('Coal_consumption')

grid.arrange(p_mc_1, ncol = 1)
```


