---
output: html_document
editor_options: 
  chunk_output_type: console
---
Lesson 4
========================================================

***

# check working directory
```{r}
getwd()
setwd("/Users/geodev/code/daan-nano/term2/EDA/lesson4/")
list.files()
```

# read libraries
```{r}
library(ggplot2)
```

# read input file
```{r}
pf <- read.csv("../lesson3/pseudo_facebook.tsv", sep = "\t")
head(pf)
```

# qplot scatterplot
```{r}
qplot(x=age, y=friend_count, data = pf)
qplot(age, friend_count, data = pf)
```


### Scatterplots and Perceived Audience Size
Notes: 

***

### Scatterplots
Notes:

```{r Scatterplots}
qplot(x=age, y=friend_count, data = pf)
qplot(age, friend_count, data = pf)
```

***

#### What are some things that you notice right away?
Response:
Vertical banding, bad data. peaks at given ages, check. large age values, check. large friend counts, check.

***

### ggplot Syntax
Notes:

```{r ggplot Syntax}
# reminder qplot syntax
qplot(x=age, y=friend_count, data = pf)
# ggplot syntax
ggplot(aes(x = age, y = friend_count + 1), data = pf) +
  # geom_point shows vertical strips as ages are reported as int but are continuous
  #geom_point(alpha = 1/20) +
  # geom_jitter shifts some points by a random value to spread out the data
  geom_jitter(alpha = 1/20) +
  xlim(min(pf$age), 90) +
  # NB. add 1 to friend_count for sqrt
  coord_trans(y = "sqrt")

# get summaries of data to fix axes
summary(pf$age)

```

***

### Overplotting
Notes: using alpha shows that most users have <1000 friends, using jitter removes much of the vertical banding but some spike remain, for instance near ~70

```{r Overplotting}

```

#### What do you notice in the plot?
Response:

***

### Coord_trans()
Notes:
the bulk of users have friend counts below 1000 and the bulk of users ares between 13 and 70 years old, the spike at ~70 is still visible of course
```{r Coord_trans()}

```

#### Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

```{r}
# ggplot syntax
ggplot(aes(x = age, y = friend_count + 1), data = pf) +
  # geom_point shows vertical strips as ages are reported as int but are continuous
  geom_point(alpha = 1/20, position = position_jitter(h = 0)) +
  xlim(min(pf$age), 90) +
  # NB. add 1 to friend_count for sqrt
  coord_trans(y = "sqrt")
```

#### What do you notice?

***

### Alpha and Jitter
friends_initiated vs age
Notes:
again some spikes at ages >60 and >90 that may be dubious
very similar distribution to age vs friend_count

```{r Alpha and Jitter}
ggplot(aes(x = age, y = friendships_initiated), data= pf) + 
  geom_point(alpha = 1/20, position = position_jitter(h = 0)) + 
  xlim(min(pf$age), 90) +
  coord_trans(y = "sqrt")
```

***

### Overplotting and Domain Knowledge
Notes:

***

### Conditional Means
Notes:

#import libraries
```{r}
install.packages('dplyr')
library('dplyr')
```

# working with dplyr
```{r}
age_groups <- group_by(pf, age)
pf.fc_by_age <- summarise(age_groups, 
          friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n())
pf.fc_by_age <- arrange(pf.fc_by_age, age)
head(pf.fc_by_age)
```

# other method for same result
```{r}
pf.fc_by_age2 <- pf %>% 
  group_by(age) %>% 
  summarise(friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n()) %>% 
  arrange(age)
head(pf.fc_by_age2)
```


```{r Conditional Means}
ggplot(aes(x = age, y = friend_count_mean), data = pf.fc_by_age2) + 
  geom_line() +
  xlim(min(pf$age), 75)
```

Create your plot!

```{r Conditional Means Plot}
ggplot(aes(x = age, y = friendships_initiated), data= pf) + 
  geom_point(alpha = 1/20, 
             position = position_jitter(h = 0),
             color = 'lightgreen') + 
  #xlim(min(pf$age), 90) +
  #coord_trans(y = "sqrt") + 
  geom_line(stat = 'summary', fun.y = mean) + 
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = .9), color = 'blue', linetype = 2) + 
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = .1), color = 'red', linetype = 2) + 
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = .5), color = 'darkgreen', linetype = 2) +
  coord_cartesian(xlim = c(min(pf$age), 90), ylim = c(0, 3000))
```

***

### Overlaying Summaries with Raw Data
Notes:

```{r Overlaying Summaries with Raw Data}

```

#### What are some of your observations of the plot?
Response:
while there are some very high friend counts, they are mostly outliers relative to the mean values, some anomalous data exist at ~70 and >75, as can be seen from the strong divergence between quartiles. The median is always lower than the mean, this is because the median is more robust than the average which is being pulled up by the extremes.

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

***

### Correlation
Notes:

```{r Correlation}
correl_ageVfriendCount <- cor.test(pf$age, pf$friend_count, method = "pearson")
correl_ageVfriendCount

# using with()
with(pf, cor.test(age, friend_count, method = 'pearson'))
```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response:
-0.02740737

***

### Correlation on Subsets
Notes:

```{r Correlation on Subsets}
# note that 'pearson' is the default and can be left out
with(subset(pf, age <= 70, select = c(age, friend_count)), cor.test(age, friend_count,
                                                                    method = 'pearson'))
with(subset(pf, age <= 70, select = c(age, friend_count)), cor.test(age, friend_count,
                                                                    method = 'spearman'))

```

***

### Correlation Methods
Notes:

***

## Create Scatterplots
Notes:

```{r}
ggplot(pf, aes(x = www_likes_received, y = likes_received)) +
  geom_point() +
  xlim(0, quantile(pf$www_likes_received, .95)) +
  ylim(0, quantile(pf$likes_received, .95)) + 
  geom_smooth(method = 'lm', color = 'green')
```

***

### Strong Correlations
Notes:

```{r Strong Correlations}

```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}
with(pf, cor.test(www_likes_received, likes_received, method = 'pearson'))
```

Response:

***

### Moira on Correlation
Notes:

***

### More Caution with Correlation
Notes:

```{r More Caution With Correlation}
install.packages('alr3')
library(alr3)
```

#check out data
```{r}
data(Mitchell)
?Mitchell
```

Create your plot!

```{r Temp vs Month}
ggplot(Mitchell, aes(x = Month, y = Temp)) + 
  geom_point()
```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.
# 0 
b. What is the actual correlation of the two variables?
(Round to the thousandths place)

```{r Noisy Scatterplots}
with(Mitchell, cor.test(Month, Temp, method = 'pearson'))
```

***

### Making Sense of Data
Notes:

```{r}
summary(Mitchell)
Mitchell
```

# new month table
```{r}
Mitchell$Year <- floor(Mitchell$Month / 12)
Mitchell$Year
Mitchell$Month
Mitchell$collapsedMonth <- (Mitchell$Month + 1) - (Mitchell$Year * 12)
head(Mitchell)
```


```{r}
ggplot(Mitchell, aes(x = collapsedMonth, y = Temp)) + 
  geom_point()
```

```{r Making Sense of Data}
ggplot(Mitchell, aes(x = collapsedMonth, y = Temp)) + 
  geom_point(aes(colour = Year)) +
  scale_x_continuous(limits = c(1, 12), breaks = seq(1,12,1))
```

***

### A New Perspective
```{r}
ggplot(Mitchell, aes(x = Month, y = Temp)) + 
  geom_point() +
  scale_x_continuous(breaks = seq(0,203,1))
```
```{r}
ggplot(aes(x=(Month%%12),y=Temp), data=Mitchell)+
  geom_point()
```


What do you notice?
Response:
The plot is still not very useful, compared to mine

Watch the solution video and check out the Instructor Notes!
Notes:

***

### Understanding Noise: Age to Age Months
Notes:

```{r Understanding Noise: Age to Age Months}
ggplot(aes(x = age, y = friend_count_mean), data = pf.fc_by_age2) + 
  geom_line()
head(pf.fc_by_age2, 10)
pf.fc_by_age2[17:19,]
```

***

### Age with Months Means

```{r Age with Months Means}
# age and dob_month variables are needed
tail(pf$dob_month, 20)
tail(pf$age, 20)
pf$age_with_months <- pf$age + (1 - pf$dob_month / 12)
pf$age_with_months
```

Programming Assignment
```{r Programming Assignment}
pf.fc_by_age_months <- pf %>% 
  group_by(age_with_months) %>% 
  summarise(friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n()) %>% 
  arrange(age_with_months)
head(pf.fc_by_age_months)
```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}
filtered_ages <- subset(pf.fc_by_age_months, age_with_months < 71)
ggplot(aes(x = age_with_months, y = friend_count_mean), 
       data = filtered_ages) + 
  geom_line()
```

```{r}
ggplot(aes(x = age_with_months, y = friend_count_mean), 
       data = subset(pf.fc_by_age_months, age_with_months < 71)) +
  geom_line()
```

***

### Smoothing Conditional Means
Notes:
```{r}
library(ggplot2)
library(gridExtra)
```

```{r Smoothing Conditional Means}
plot1 <- ggplot(aes(x = age, y = friend_count_mean), 
                data = subset(pf.fc_by_age2, age < 71)) + 
  geom_line() +
  geom_smooth()

plot2 <- ggplot(aes(x = age_with_months, y = friend_count_mean), 
       data = subset(pf.fc_by_age_months, age_with_months < 71)) +
  geom_line() +
  geom_smooth()

plot3 <- ggplot(aes(x = round(age / 5) * 5, y = friend_count), 
                data = subset(pf, age < 71)) + 
  geom_line(stat = 'summary', fun.y = mean)

grid.arrange(plot1, plot2, plot3, ncol = 1)
```

***

### Which Plot to Choose?
Notes:

***

### Analyzing Two Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

