---
title: "Two variables investigation"
output: html_document
editor_options: 
  chunk_output_type: console
---
# set working directory
```{r}
getwd()
setwd("/Users/geodev/code/daan-nano/term2/EDA/lesson4")
list.files()
```


# import packages
```{r}

```

# load libraries
```{r}
library(ggplot2)
library(gridExtra)
```

# read in data
```{r}
data(diamonds)
summary(diamonds)
names(diamonds)
str(diamonds)
```

# first plot
```{r}
p1 <- ggplot(aes(x = price, y = x), data = diamonds) +
  geom_point()
p2 <- ggplot(aes(x = price, y = y), data = diamonds) +
  geom_point()
p3 <- ggplot(aes(x = price, y = z), data = diamonds) +
  geom_point()

grid.arrange(p1, p2, p3, ncol = 1)
```

# observations
# there appears to be a positive correlation between carat and price, but the spread of price is largest for diamonds of 1 to 2 carat, above and below, the range of possible prices is much smaller. There are clear bandas of prices for specific carats, suggesting some grouping of carats whereas a continous distribution would be expected for mineral grain sizes. The data appeared to be cutoff at or near a maximum. The largest carat is not the most expensive diamond, so carat is not the only variable defining price. Outliers, exponential distribution.

# Correlations
```{r}
price_x_correl <- with(diamonds, cor.test(price, x, method = 'pearson'))
price_y_correl <- with(diamonds, cor.test(price, y, method = 'pearson'))
price_z_correl <- with(diamonds, cor.test(price, z, method = 'pearson'))
price_x_correl
price_y_correl
price_z_correl
```

# price vs depth
```{r}
ggplot(aes(x = depth, y = price), data = diamonds) + 
  geom_point(alpha = .01) +
  scale_x_continuous(breaks = seq(50, 70, 2))
```

# price vs depth correlation
```{r}
price_depth_correl <- with(diamonds, cor.test(price, depth, method = 'pearson'))
price_depth_correl
```

# scatterplot of 99 percentile price vs 99 percentile carat
```{r}
ggplot(aes(x = carat, y = price), data = diamonds) + 
  geom_point() +
  xlim(0, quantile(diamonds$carat, .99)) +
  ylim(0, quantile(diamonds$price, .99))
```

# Create a new variable for volume in the diamonds data frame (very rough approximation for a diamond's volume)
```{r}
diamonds$volume <- diamonds$x * diamonds$y * diamonds$z
head(diamonds$volume)
```

# Create a scatterplot of price vs. volume (x * y * z)
```{r}
ggplot(aes(x = volume, y = price), data = diamonds) + 
  geom_point()
```
# observations:
# there are several outliers, the overall trend is consistent but some point fall on the same trend with a different slope

# count of zero volume points
```{r}
count_zero_volume <- length(which(diamonds$volume == 0))
count_zero_volume
```

# volume filtered diamonds
```{r}
filtered_volumes <- subset(diamonds, diamonds$volume > 0 & diamonds$volume < 800)
summary(filtered_volumes)
```

# plof of filtered diamonds
```{r}
ggplot(aes(x = volume, y = price), data = filtered_volumes) + 
  geom_point()
```


# correlation of price vs filtered volume
```{r}
price_filtVolume_correl <- with(filtered_volumes, cor.test(price, volume, method = 'pearson'))
price_filtVolume_correl
```

# Subset the data and add linear model
```{r}
ggplot(aes(x = volume, y = price), data = filtered_volumes) +
  geom_point(alpha = .05) +
  geom_smooth()
```
# comments
# neither a linear nor the default model are a good match to the data, they should not be used for price prediction.

# diamondsByClarity
```{r}
clarity_groups <- group_by(diamonds, clarity)
diamondsByClarity <- summarise(clarity_groups,
                               mean_price = mean(price),
                               median_price = median(price),
                               min_price = min(price),
                               max_price = max(price),
                               n = n())
diamondsByClarity <- arrange(diamondsByClarity, clarity)
head(diamondsByClarity)
```

# bar charts of mean prices
```{r}
diamonds_by_clarity <- group_by(diamonds, clarity)
diamonds_mp_by_clarity <- summarise(diamonds_by_clarity, mean_price = mean(price))

diamonds_by_color <- group_by(diamonds, color)
diamonds_mp_by_color <- summarise(diamonds_by_color, mean_price = mean(price))
```

# bar charts
```{r}
p_mp_1 <- ggplot(aes(x = mean_price), data = diamonds_mp_by_clarity) + 
  geom_bar()
p_mp_2 <- ggplot(aes(x = mean_price), data = diamonds_mp_by_color) + 
  geom_bar()

grid.arrange(p_mp_1, p_mp_2, ncol = 1)
```

