---
output: html_document
editor_options: 
  chunk_output_type: console
---
Lesson 5
========================================================

# check working directory
```{r}
getwd()
setwd("/Users/geodev/code/daan-nano/term2/EDA/lesson5/")
list.files()
```

# import packages
```{r}
install.packages("tidyr")
```
```{r}
install.packages("dplyr")
```

# read libraries
```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(readxl)
library(gridExtra)
library(grid)
```

# read data into dataframe
```{r}
# define path name
pf <- read.csv('../lesson3/pseudo_facebook.tsv', sep = '\t')
head(pf)
str(pf)
```

# Friend counts
```{r}
pf.fc_by_age_gender <- pf %>% 
  filter(!is.na(gender)) %>% 
  group_by(age, gender) %>% 
  summarise(mean_friend_count = mean(friend_count),
            median_friend_count = median(friend_count),
            n = n()) %>% 
  ungroup() %>% 
  arrange(age)

head(pf.fc_by_age_gender)
```

```{r}
ggplot(data = pf.fc_by_age_gender, aes(x = age, y = mean_friend_count, color = gender)) +
  geom_line()
```


### Multivariate Data
Notes:

***

### Moira Perceived Audience Size Colored by Age
Notes:

***

### Third Qualitative Variable
Notes:

```{r Third Qualitative Variable}
ggplot(aes(x = gender, y = age),
       data = subset(pf, !is.na(gender))) + geom_histogram()
```

***

### Plotting Conditional Summaries
Notes:

```{r Plotting Conditional Summaries}

```

***

### Thinking in Ratios
Notes:

***

### Wide and Long Format
Notes:

***

### Reshaping Data
Notes:

```{r}
install.packages('reshape2')
library(reshape2)
```

```{r}
pf.fc_by_age_gender.wide <- dcast(pf.fc_by_age_gender, 
                                  age ~ gender, 
                                  value.var = 'median_friend_count')
head(pf.fc_by_age_gender.wide)
```


***

### Ratio Plot
Notes:

```{r Ratio Plot}
# create ratio value
pf.fc_by_age_gender.wide$ratio <- pf.fc_by_age_gender.wide$female / pf.fc_by_age_gender.wide$male

head(pf.fc_by_age_gender.wide)

pf.fc_by_age_gender.wide %>% 
  ggplot(aes(x = age, y =  ratio)) +
  geom_hline(yintercept = 1, linetype = 2, alpha = .3) +
  geom_line()

# same without external ratio variable
pf.fc_by_age_gender.wide %>% 
  ggplot(aes(x = age, y =  female/male)) +
  geom_hline(yintercept = 1, linetype = 2, alpha = .3) +
  geom_line()


```

***

### Third Quantitative Variable
Notes:

```{r Third Quantitative Variable}
pf$year_joined <- floor(2014 - pf$tenure / 365)
tail(pf$year_joined)
summary(pf$year_joined)
table(pf$year_joined)
```

***

### Cut a Variable
Notes:

```{r Cut a Variable}
pf$year_joined.bucket <- cut(pf$year_joined, c(2004, 2009, 2011, 2012, 2014))
table(pf$year_joined.bucket)  
```

***

### Plotting it All Together
Notes:

```{r Plotting it All Together}
ggplot(aes(x = age, y = friend_count),
       data = subset(pf, !is.na(year_joined.bucket))) +
  geom_line(aes(color = year_joined.bucket), stat = 'summary', fun.y = mean) + 
  # plotting the grand mean (i.e. the mean of the means)
  geom_line(aes(), stat = 'summary', fun.y = mean, linetype = 2)
```

***

### Plot the Grand Mean
Notes:

```{r Plot the Grand Mean}

```

***

### Friending Rate
Notes:

```{r Friending Rate}
# count of friends by day since starting using service for users with day >= 1
with(subset(pf, tenure >= 1), summary(friend_count / tenure))
```

***

### Friendships Initiated
Notes:

What is the median friend rate?
0.22
What is the maximum friend rate?
417
```{r Friendships Initiated}
ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) + 
  geom_line(aes(color=year_joined.bucket), 
            stat = 'summary',
            fun.y = mean)
```

# prop_initiated: proportion of friendships that the user initiated.
```{r}
str(pf)
# subset friends with > 0 friends and > 0 intiated
pf.social <- subset(pf, pf$friend_count > 0 & pf$friendships_initiated > 0)
pf.social$prop_initiated <- pf.social$friend_count / pf.social$friendships_initiated
head(pf.social$prop_initiated)

#other method with ifelse
pf$prop_initiated <- ifelse(pf$friend_count > 0, pf$friendships_initiated/pf$friend_count, 0)
```

```{r}
pi1 <- ggplot(data = pf, aes(x = tenure, y = prop_initiated)) + 
  geom_line(aes(colour = year_joined.bucket), 
            stat = 'summary',
            fun.y = median)

pi2 <- ggplot(data = pf, aes(x = 30 * round(tenure/30), y = prop_initiated)) + 
  geom_line(aes(colour = year_joined.bucket), 
            stat = 'summary',
            fun.y = median)

grid.arrange(pi1, pi2, ncol=1)
```

# mean of largest prop_init group
```{r}
# subset friends w > 0 friends & group (2012,2014]
pf.social.group4 <- subset(pf.social, pf.social$year_joined.bucket == '(2012,2014]')
pf.social.group4$prop_initiated2 <- pf.social.group4$friendships_initiated / pf.social.group4$friend_count
mean(pf.social.group4$prop_initiated2)

```


***

### Bias-Variance Tradeoff Revisited
Notes:

```{r Bias-Variance Tradeoff Revisited}

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

```

```{r}
ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  #geom_line(aes(color = year_joined.bucket), stat = "summary", fun.y = mean)
  geom_smooth(aes(color = year_joined.bucket))
```

***

### Sean's NFL Fan Sentiment Study
Notes:

***

### Introducing the Yogurt Data Set
Notes:

***

### Histograms Revisited
Notes:

```{r Histograms Revisited}
list.files()
yo <- read.csv('yogurt.csv')
head(yo)
str(yo)

# change the id from int to factor
yo$id <- factor(yo$id)
str(yo)

ggplot(data = yo, aes(x = price)) +
  geom_histogram(binwidth=10)
summary(yo$price)
summary(yo)
unique(yo$price)
table(yo$price)
```

***

### Number of Purchases
Notes:

```{r Number of Purchases}
yo <- transform(yo, all.purchases = strawberry + blueberry + pina.colada + plain + mixed.berry)
head(yo$all.purchases, 20)
summary(yo$all.purchases)

#alternate verbose form
yo$all.purchases <- yo$strawberry + yo$blueberry + yo$pina.colada + yo$plain + yo$mixed.berry

ggplot(data = yo, aes(x = all.purchases)) + 
  geom_histogram(binwidth = 1)
```

***

### Prices over Time
Notes:

```{r Prices over Time}
ggplot(data = yo, aes(x = time, y = price)) + 
  geom_point(alpha = .05)
```

***

### Sampling Observations
Notes:

***

### Looking at Samples of Households

```{r Looking at Sample of Households}
set.seed(1977)
sample.ids <- sample(levels(yo$id), 16)
sample.ids

ggplot(aes(x = time, y = price), 
       data = subset(yo, id %in% sample.ids)) +
  facet_wrap( ~ id) +
  geom_line() +
  geom_point(aes(size = all.purchases), pch = 1)
```

***

### The Limits of Cross Sectional Data
Notes:

***

### Many Variables
Notes:

***

### Scatterplot Matrix
Notes:
# install packages and load
```{r}
install.packages('GGally')
library('GGally')
```
# matrix
```{r}
theme_set(theme_minimal(20))
```
# set seed
```{r}
set.seed(1836)
pf_subset <- pf[, c(2:15)]
names(pf_subset)
ggpairs(pf_subset[sample.int(nrow(pf_subset), 1000), ])

```



***

### Even More Variables
Notes:
```{r}
list.files()
```

***

### Heat Maps
Notes:

```{r}
nci <- read.table("nci.tsv")
colnames(nci) <- c(1:64)
```

```{r}
library(reshape2)
nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

ggplot(aes(y = gene, x = case, fill = value),
  data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))
```


***

### Analyzing Three of More Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

