# Wine Quality Exploration by Robert Leckenby

========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE}
# check working directory
getwd()
setwd("/Users/geodev/code/daan-nano/ExplAndSummData_project")
list.files()
```


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# install packages
install.packages("ggplot2", dependencies = T) 
install.packages("knitr", dependencies = T)
install.packages("dplyr", dependencies = T)

# load libraries
library(ggplot2)
library(knitr)
library(dplyr)
library(gridExtra)
library(purrr)
library(tidyr)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
rwine_raw <- read.csv('wineQualityReds.csv')
```

========================================================

# General Information about data source and attributes

Citation Request:
  This dataset is public available for research. The details are described in
  [Cortez et al., 2009]. 
  Please include this citation if you plan to use this database:

  P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
  Modeling wine preferences by data mining from physicochemical properties.
  In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

  Available at: [@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016
                [Pre-press (pdf)] 
                  http://www3.dsi.uminho.pt/pcortez/winequality09.pdf
                [bib] http://www3.dsi.uminho.pt/pcortez/dss09.bib

1. Title: Wine Quality 

2. Sources
   Created by: Paulo Cortez (Univ. Minho), Antonio Cerdeira, Fernando Almeida, 
   Telmo Matos and Jose Reis (CVRVV) @ 2009
   
3. Past Usage:

  P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
  Modeling wine preferences by data mining from physicochemical properties.
  In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

  In the above reference, two datasets were created, using red and white wine
  samples.
  The inputs include objective tests (e.g. PH values) and the output is based
  on sensory data (median of at least 3 evaluations made by wine experts). Each 
  expert graded the wine quality between 0 (very bad) and 10 (very excellent). 
  Several data mining methods were applied to model these datasets under a 
  regression approach. The support vector machine model achieved the best
  results. Several metrics were computed: MAD, confusion matrix for a fixed 
  error tolerance (T), etc. Also, we plot the relative importances of the input 
  variables (as measured by a sensitivity analysis procedure).
 
4. Relevant Information:

   The two datasets are related to red and white variants of the Portuguese 
   "Vinho Verde" wine.
   For more details, consult: http://www.vinhoverde.pt/en/ or the reference 
   [Cortez et al., 2009].
   Due to privacy and logistic issues, only physicochemical (inputs) and sensory
   (the output) variables are available (e.g. there is no data about grape 
   types, wine brand, wine selling price, etc.).

   These datasets can be viewed as classification or regression tasks.
   The classes are ordered and not balanced (e.g. there are munch more normal 
   wines than excellent or poor ones). Outlier detection algorithms could be
   used to detect the few excellent or poor wines. Also, we are not sure if all 
   input variables are relevant. So it could be interesting to test feature 
   selection methods. 

5. Number of Instances: red wine - 1599; white wine - 4898. 

6. Number of Attributes: 11 + output attribute
  
   Note: several of the attributes may be correlated, thus it makes sense to 
   apply some sort of feature selection.

7. Attribute information:

   For more information, read [Cortez et al., 2009].

   Input variables (based on physicochemical tests):
   1 - fixed acidity (tartaric acid - g / dm^3)
   2 - volatile acidity (acetic acid - g / dm^3)
   3 - citric acid (g / dm^3)
   4 - residual sugar (g / dm^3)
   5 - chlorides (sodium chloride - g / dm^3
   6 - free sulfur dioxide (mg / dm^3)
   7 - total sulfur dioxide (mg / dm^3)
   8 - density (g / cm^3)
   9 - pH
   10 - sulphates (potassium sulphate - g / dm3)
   11 - alcohol (% by volume)
   Output variable (based on sensory data): 
   12 - quality (score between 0 and 10)

8. Missing Attribute Values: None

9. Description of attributes:

   1 - fixed acidity: most acids involved with wine or fixed or nonvolatile (do 
   not evaporate readily)
   2 - volatile acidity: the amount of acetic acid in wine, which at too high of
   levels can lead to an unpleasant, vinegar taste
   3 - citric acid: found in small quantities, citric acid can add 'freshness' 
   and flavor to wines
   4 - residual sugar: the amount of sugar remaining after fermentation stops, 
   it's rare to find wines with less than 1 gram/liter and wines with greater 
   than 45 grams/liter are considered sweet
   5 - chlorides: the amount of salt in the wine
   6 - free sulfur dioxide: the free form of SO2 exists in equilibrium between 
   molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial 
   growth and the oxidation of wine
   7 - total sulfur dioxide: amount of free and bound forms of S02; in low
   concentrations, SO2 is mostly undetectable in wine, but at free SO2 
   concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine
   8 - density: the density of water is close to that of water depending on the 
   percent alcohol and sugar content
   9 - pH: describes how acidic or basic a wine is on a scale from 0 (very 
   acidic) to 14 (very basic); most wines are between 3-4 on the pH scale
   10 - sulphates: a wine additive which can contribute to sulfur dioxide gas
   (S02) levels, wich acts as an antimicrobial and antioxidant
   11 - alcohol: the percent alcohol content of the wine
   Output variable (based on sensory data): 
   12 - quality (score between 0 and 10)
   
========================================================

# Univariate Plots Section

> **Tip**: In this section, you should perform some preliminary exploration of
your dataset. Run some summaries of the data and create univariate plots to
understand the structure of the individual variables in your dataset. Don't
forget to add a comment after each plot or closely-related group of plots!
There should be multiple code chunks and text sections; the first one below is
just to help you get started.

```{r echo=FALSE, Univariate_Plots}
summary(rwine_raw)
names(rwine_raw)
str(rwine_raw)
```

From str(), we see that the data set consists of 1599 observations, each with 13 variables.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality), data = rwine_raw) +
  geom_bar() + 
  scale_x_continuous(limits = c(0,10), breaks = seq(0,10,1))
```

```{r echo=False, message=FALSE, warning=FALSE}
summary(rwine_raw$quality)
```

A quick look at a bar chart of wine qualities shows us a near normal 
distribution with a mean of ~5.6 and a median value of 6, whereas the min and 
max are 3 and 8, respectively. No wines were given a 0 nor a 10, but there is a 
bias towards higher quality wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol), data = rwine_raw) +
  geom_histogram()
```

```{r echo=False, message=FALSE, warning=FALSE}
summary(rwine_raw$alcohol)
```

Looking at a histogram of alchohol content shows a different story than wine 
qualities: here we see a left-skewed distribution of alcohol contents with a 
median of 10.2 and a mean of 10.42, but with a long high value tail and a
maximum -which appears to be an outlier- at 14.90

Here let us note that there is only one categorical variable in the data set:
the wine quality, so it will be interesting to look at this against multiple 
variables later on.

Now we have had a quick look at quality and alcohol content, let's look at some
other attributes that are expected to have an impact on quality and health, but
without comparing them to eachother for the moment


```{r echo=FALSE, message=FALSE, warning=FALSE}
# Acid and acidity
p1 <- ggplot(aes(x = fixed.acidity), data = rwine_raw) +
  geom_histogram(binwidth = 0.2)
p2 <- ggplot(aes(x = volatile.acidity), data = rwine_raw) +
  geom_histogram(binwidth = 0.02)
p3 <- ggplot(aes(x = citric.acid), data = rwine_raw) +
  geom_histogram(binwidth = 0.02)
grid.arrange(p1, p2, p3, ncol=1)
```

Fixed and volatile acidity show similar histogram characterisitcs, namely a 
close to normal distribution with a long high value tail and some probable 
outliers. Changing the binwidths for all three variables shows more 
heterogeneity in the data than initially observed with the default bin number
value (30), and so it will be important to take this into account when looking 
at multivariate relations.

Citric acid, as well as some peaks at some values (0, ~0.25, ~0.49), shows quite
a large variation in counts.

```{r echo=FALSE, message=FALSE, warning=FALSE}
sum(rwine_raw$citric.acid == 1)
```

As seen above, there is a single outlier with a citric acid content of 1. It
will be interesting to investigate whether this outlier is a true data point or
a measurement error for example.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# plotting histograms of all variables at once to get quick look overview
rwine_raw %>%
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(value)) +
    facet_wrap(~ key, scales = "free") +
    geom_histogram()
```

Having a look at all variables at once - but ignoring those discussed already -
we see that chlorides show a normal distribution with a very long tail, that 
density and pH are normally distributed at this scale, that sulfur values (both
free and total) show a skewed distribution with long tails and that sugars and 
sulphates might be close to normally distributed, but with asymetric tails.

Some properties appear to have significant outliers which may have to be ignored
or at least considered in any analysis. These are:
  - chlorides
  - free.sulphur.dioxide
  - residual.sugar
  - sulphates
  - total.sulfur.dioxide
  - volatile.acidity
  
## Outliers  
Let us know look at these outliers more closely:

```{r echo=FALSE, message=FALSE, warning=FALSE}
p.chlorides <- ggplot(aes(chlorides), data = rwine_raw) +
  geom_histogram() +
  scale_y_continuous(limits = c(0,5), breaks = seq(0,5, 1))
p.free.sulfur.dioxide <- ggplot(aes(free.sulfur.dioxide), data = rwine_raw) +
  geom_histogram() +
  scale_y_continuous(limits = c(0,5), breaks = seq(0,5, 1))
p.residual.sugar <- ggplot(aes(residual.sugar), data = rwine_raw) +
  geom_histogram() +
  scale_y_continuous(limits = c(0,5), breaks = seq(0,5, 1))
p.sulphates <- ggplot(aes(sulphates), data = rwine_raw) +
  geom_histogram() +
  scale_y_continuous(limits = c(0,5), breaks = seq(0,5, 1))
p.total.sulfur.dioxide <- ggplot(aes(total.sulfur.dioxide), data = rwine_raw) +
  geom_histogram() +
  scale_y_continuous(limits = c(0,5), breaks = seq(0,5, 1))
p.volatile.acidity <- ggplot(aes(volatile.acidity), data = rwine_raw) +
  geom_histogram() +
  scale_y_continuous(limits = c(0,5), breaks = seq(0,5, 1))

grid.arrange(p.chlorides, p.free.sulfur.dioxide, p.residual.sugar, p.sulphates,
             p.total.sulfur.dioxide, p.volatile.acidity, ncol=2)
```

Plotting those variables that appear to have outliers, but limiting the counts
to 5 to see if there is litterally only 1 outlier in each case, shows a more 
nuanced picture, with some intermediary values. This suggests that we are not
dealing with measurement error on a single value, but no conclusions can be 
drawn at this point.
The result of this quick look is that I will not exclude these points from the 
analysis.

```{r echo=FALSE, message=False, warning=FALSE}
str(rwine_raw)
```



# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!