---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Lesson 3

-----

# read in data

```{r}
getwd()
list.files()
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')
names(pf)
```

# histogram of birthdays
```{r}
install.packages('ggplot2')
library(ggplot2)

#get column names
names(pf)
summary(pf)
str(pf)
# use qplot to see data
qplot(x = dob_day, data = pf) +
  scale_x_continuous(breaks=1:31) +
  facet_wrap(~dob_month, ncol=3)
```

# histogram of friend count
```{r}
names(pf)
#qplot(x = friend_count, data = pf, xlim = c(0,1000))

#alternative method to change axis
qplot(x = friend_count, data = subset(pf, !is.na(gender)), binwidth = 25) +
  scale_x_continuous(limits = c(0,1000), breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender,ncol = 2)
```

#stats by gender
```{r}
table(pf$gender)
by(pf$friend_count, pf$gender, summary)
```

# Tenure
```{r}
qplot(x = tenure, data = pf, binwidth = 30,
      color = I('black'),fill = I('#099DD9'))
```
# Tenure in years
```{r}
str(pf)
qplot(x = tenure/365, data = pf, binwidth = 1,
      xlab="number of years wasted on facebook",
      ylab="number of fools wasting time on facebook",
      color = I('black'),fill = I('#099DD9'))
```
# alternative tenure in years
```{r}
ggplot(aes(x = tenure/365), data = pf,) +
  geom_histogram(binwidth = .25, color = 'black', fill = '#F79420') +
  scale_x_continuous(limits = c(0,7), breaks = seq(0, 7, 1))
```

# ages investigation
```{r}
ggplot(aes(x = age), data = pf) +
  geom_histogram(binwidth = 1, color = 'black', fill = 'green') +
  scale_x_continuous(limits = c(0,120), breaks = seq(0,120,5)) + 
  labs(x = "Age (years)", y = "nb of users")
```

# Transforming data
```{r}
qplot(x = friend_count, data = pf)
summary(pf$friend_count)
summary(log10(pf$friend_count + 1))
summary(sqrt(pf$friend_count))
```

# Transformating data and plotting multiple plots
```{r}
install.packages("gridExtra")
library(gridExtra)
```

```{r}
# define individual plots
p1 = ggplot(aes(x = friend_count), data = pf) +
  geom_histogram(binwidth = 1, color = 'black', fill = 'red') +
  scale_x_continuous(limits = c(0,5000), breaks = seq(0,5000,500)) #+ 
  #labs(x = "friend_count linear", y = "nb of users")
p2 = ggplot(aes(x = log10(friend_count + 1)), data = pf) +
  geom_histogram(color = 'black', fill = 'green') +
  scale_x_continuous(limits = c(0,5)) #+ 
  #labs(x = "friend_count", y = "log10 nb of users")
p3 = ggplot(aes(x = sqrt(friend_count)), data = pf) +
  geom_histogram(binwidth = 1, color = 'black', fill = 'blue') +
  scale_x_continuous(limits = c(0,120), breaks = seq(0,120,5)) #+ 
  #labs(x = "sqrt friend_count", y = "nb of users")
# arrange plots in grid
grid.arrange(p1, p2, p3, ncol=1)
```

# Solution
```{r}
p1 <- qplot(x = friend_count, data = pf)
p2 <- qplot(x = log10(friend_count + 1), data = pf)
p3 <- qplot(x = sqrt(friend_count), data = pf)

grid.arrange(p1, p2, p3, ncol=1)
```

# Solution 2
```{r}
p1 <- ggplot(aes(x = friend_count), data = pf) + geom_histogram()
p2 <- p1 + scale_x_log10()
p3 <- p1 + scale_x_sqrt()
grid.arrange(p1, p2, p3, ncol=1)
```

# scaling layer solution
```{r}
logScale <- ggplot(aes(x = log10(friend_count + 1)), data = pf) + geom_histogram()
countScale <- ggplot(aes(x = friend_count), data = pf) +
  geom_histogram() +
  scale_x_log10()
grid.arrange(logScale, countScale, ncol=2)
```

# frequency polygons starting histograms
```{r}
qplot(x = friend_count, data = subset(pf, !is.na(gender)), binwidth=10) +
  scale_x_continuous(limits = c(0,1000), breaks = seq(0,1000,50)) +
  facet_wrap(~gender)
```

# frequency polygons
```{r}
qplot(x = friend_count, data = subset(pf, !is.na(gender)),
      binwidth=10,
      geom = 'freqpoly',
      color = gender) +
  scale_x_continuous(limits = c(0,1000), breaks = seq(0,1000,50))
```

# frequency polygons with changed y-axis
```{r}
qplot(x = friend_count, y = ..count../sum(..count..),
      data = subset(pf, !is.na(gender)),
      xlab = 'Friend count',
      ylab = 'Proportion of users with given friend count',
      binwidth=10,
      geom = 'freqpoly',
      color = gender) +
  scale_x_continuous(limits = c(0,1000), breaks = seq(0,1000,50))
```

# frequency polygons of www_likes
```{r}
qplot(x = www_likes, y = ..count../sum(..count..),
      data = subset(pf, !is.na(gender)),
      xlab = 'www_likes',
      ylab = 'Proportion of users with given www_likes',
      geom = 'freqpoly',
      color = gender) +
  scale_x_continuous() +
  scale_x_log10()
```

# count of likes for m and f
```{r}
by(pf$www_likes, pf$gender, sum)
```

# Boxplot
```{r}
qplot(x = gender, y = friend_count, 
      data = subset(pf, !is.na(gender)), 
      geom = 'boxplot') +
  scale_y_continuous(limits = c(0,1000), breaks = seq(0,1000,50)) #+
  #scale_y_log10()
```

# Boxplot solution 1 !! ylim trims data and so changes boxplot
```{r}
qplot(x = gender, y = friend_count, 
      data = subset(pf, !is.na(gender)), 
      geom = 'boxplot', ylim = c(0,1000))
```

# Boxplot solution 2 !! as above
```{r}
qplot(x = gender, y = friend_count, 
      data = subset(pf, !is.na(gender)), 
      geom = 'boxplot') +
  scale_y_continuous(limits = c(0,1000))
```

# Boxplot solution 3 - no trim to data
```{r}
qplot(x = gender, y = friend_count, 
      data = subset(pf, !is.na(gender)), 
      geom = 'boxplot') +
  coord_cartesian(ylim = c(0,1000))
```
